Introductory statistics with `intRo`

### 2016-05-26 jcgs-paper-2016/paper.Rmd
Start restructuring and removing a lot

---
title: "Introductory statistics with `intRo`"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
fontsize: 12pt
abstract: |
  `intRo` is a web-based application for performing basic data analysis and statistical routines. Leveraging the power of R and Shiny, `intRo` implements common statistical functions in an extensible modular structure, while including a point-and-click interface for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as Excel and JMP. `intRo` is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying `intRo`, as well as highlight some challenges and advantages of reactive programming.
keywords: Graphical User Interface, Interactivity, Programming Paradigms, Reactive Programming, Reproducibility, Statistical Software
bibliography: refs.bib
blind: false
---

@@

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the user's R script. By default, the output will include the R code, but if the user elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results.

@@

Each module is a self contained set of R code that is dynamically added to the application at run time. What this means is that `intRo` can be easily extended by the addition of modules within the framework underlying the application. This extensibility allows for `intRo` to be customized for an individual classroom setting, tailored to the needs of a particular course.

As an example of possible customization, an instructor may wish to introduce hypothesis testing from a nonparametric framework rather than distributional. In this case, a wilcoxon rank sum test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork `intRo` from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `libraries.R`, `observe.R`, `output.R`, `reactive.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a Shiny application. Further details on the creation of new modules can be found in section \ref{modules}.

@@

The process by which `intRo` is dynamically created upon running the application through module sourcing is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the navigation file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

# Using `intRo`

Although users can access intRo from \url{http://www.intro-stats.com}, course instructors may wish to customize their own instance.

## `intRo` package

`intRo` can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package `intRo`. Currently, the package is only available on GitHub, and can be installed using the devtools package as follows:

If the intent is to use a specific instance of `intRo` where many users will access it at the same time, such as in an introductory statistics class, it may be preferable to deploy a custom instance of `intRo` to a publicly accessible URL. The package provides a function `deploy_intRo` which is a wrapper for the `deployApp` function contained in the shinyapps package. Once the shinyapps package is installed and configured, `deploy_intRo` will upload `intRo` as an application on the user's account. The function takes the same arguments as `run_intRo`, so it can be deployed with a custom selection of modules, and a customized theme. It also takes an additional argument `google_analytics`, which allows the specification of a Google Analytics tracking ID. It also takes `...` as additional arguments to be passed into the `deployApp` routine. For example, if we wished to deploy the instance of `intRo` we ran previously, we would call it like so:

## In the classroom

`intRo`'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as JMP and Excel, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. JMP, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using R itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like rmarkdown, but this again is a bit of a learning curve for the new statistics student).

We believe that `intRo` targets these key limitations. The interface is designed to be quick, seamless, and easy to understand. Because `intRo` is web-based, the statistical analysis can be performed from nearly any modern device. R underlies the analysis, but users need not be familiar with R or programming to perform it. The focal point of the interface is the output panel, allowing the student to remain focused on the results and limiting the need to fight with the interface itself to get work done. For those students who may have a natural inclination for programming, executable code is generated to allow a local reproduction of the analysis.

The features of `intRo` were selected in order to apply to the most commonly used lesson plans in early statistics courses. The textbook "The Art and Science of Learning from Data" by Agresti and Franklin [-@agresti2012statistics] references many of the focal points of introductory statistics, and `intRo` supports performing a majority of them. Part one, titled "Gathering and Exploring Data", covers graphical summaries, measures of spread and center, and avoiding the misuse of graphics in situations they are not appropriate. `intRo` includes modules for both graphical and numerical summaries covering these topics. Furthermore, `intRo` limits the display of variables to ones that are appropriate for a chosen plot type, helping to reinforce the appropriateness of particular graphics (Table \ref{tab:plottypes}).

\begin{table*}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\caption{The plot and variable types allowed in `intRo`'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\label{tab:plottypes}
\end{table*}

Part two of this text discusses probability and sampling distributions, which is outside the current realm of `intRo`, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in `intRo`. Finally, part four of the text discusses the association between variables, including regression, functions that `intRo` performs with ease. Some of the more advanced topics of this section are currently unsupported by `intRo`, including multiple regression.

Statistics courses with a lab component can most benefit from the use of `intRo`. As students learn the aforementioned concepts, they can practice using the tools to perform the statistical tests and summaries. They simply upload a dataset and click the module appropriate to the section of the class. Once they choose the options desired by the instructor (variables, confidence levels, etc.), they can view the results and ultimately print the output for turning in. Early labs can walk the user through the limited number of button presses and clicks necessary, while later labs can more vaguely describe the intended output in hopes that the user can easily navigate the `intRo` interface to obtain the results. For topics currently unsupported by `intRo`, the instructor can supplement its use with other web-based modules that are publicly available or create their own modules within `intRo` as detailed in section \ref{modules}.

To fully understand `intRo`, it is important to expand upon three underlying concepts. The first is the idea of modularity in the context of a Shiny application. The second is the idea of reproducibility in the context of a graphical user interface. Finally, the third is the idea of reactive programming in the context of a web application.

@@

This script is most immediately similar to standard R code. In this case, a function `cont.table` is created which, depending on the values of the parameters, ultimately returns a contingency table. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate`. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`. More details on the reasons and choices behind `interpolate` can be found in section \ref{reproducibility}.

@@

## Reproducibility
Though they can make common actions quicker and interacting with software much simplier, many Graphical User Interfaces for scientific software packages come with a major drawback. That is, the actions taken by the user are typically not reproducible. Code scripting, on the other hand, allows us to share a file containing code, and another user can execute the code to reproduce the same results. In terms of the scientific process, where results can and should be scrutinized by other parties, the utility of GUIs are greatly limited.

We have designed `intRo` modules to reproduce actions taken by the user so that they can obtain R code associated with each module interacted with. When a user chooses to store the results associated with a particular module, R code yielding the visible results is automatically added to the code window at the bottom. This framework yields several advantages:

We designed the modules in such a way that the code printed to the console is the same as the code executed by the server, which makes the process of writing an `intRo` module significantly more seamless. The workhorse function for this reproducibility framework is called `interpolate`, which is based on a function proposed by Hadley Wickham [-@interpolate]. The idea of interpolation is that by using non-standard evaluation, the call can be captured, printed, and evaluated, with all the appropriate arguments substituted into the call automatically. This means that depending on the options selected in `intRo`, the corresponding arguments to functions are appropriately set for reproduction in a standard R session.

@@

@@

# Supplementary material
All code and documents related to this manuscript are available at \url{https://github.com/gammarama/intRo}.

>>>>

---
title: "Introductory statistics with `intRo`"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
fontsize: 12pt
abstract: |
  `intRo` is a web-based application for performing basic data analysis and statistical routines. Leveraging the power of R and Shiny, `intRo` implements common statistical functions in an extensible modular structure, while including a point-and-click interface for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as Excel and JMP. `intRo` is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying `intRo`, as well as highlight some challenges and advantages of reactive programming.
keywords: Interactivity, Programming Paradigms, Reactive Programming, Reproducibility, Statistical Software
bibliography: refs.bib
blind: false
---

@@

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the user's R script. By default, the output will include the R code, but if the user elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results. This reproducibility framework has a number of key advantages:

@@

To fully understand `intRo`, it is important to expand upon two underlying concepts. The first is the idea of modularity in the context of a Shiny application. The second is the idea of reactive programming in the context of a web application.

@@

`intRo`'s user interface and functionality is dynamically generated depending on the set of modules enabled. This process is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the navigation file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

@@

This script is most immediately similar to standard R code. In this case, a function `cont.table` is created which, depending on the values of the parameters, ultimately returns a contingency table. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate` [@interpolate]. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`.

@@

@@

# Using and Deploying `intRo`

Although users can access intRo from \url{http://www.intro-stats.com}, course instructors may wish to download, customize, and their own instance, perhaps with new modules or modified theming or functionality. `intRo` can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package `intRo`. Currently, the package is only available on GitHub, and can be installed using the devtools package as follows:

If the intent is to use a specific instance of `intRo` where many users will access it at the same time, such as in an introductory statistics class, it may be preferable to deploy a custom instance of `intRo` to a publicly accessible URL. The package provides a function `deploy_intRo` which is a wrapper for the `deployApp` function contained in the shinyapps package. Once the shinyapps package is installed and configured, `deploy_intRo` will upload `intRo` as an application on the user's ShinyApps.io account. The function takes the same arguments as `run_intRo`, so it can be deployed with a custom selection of modules, and a customized theme. It also takes an additional argument `google_analytics`, which allows the specification of a Google Analytics tracking ID. It also takes `...` as additional arguments to be passed into the `deployApp` routine. For example, if we wished to deploy the instance of `intRo` we ran previously, we would call it like so:

@@

# Supplementary material

All code and documents related to this manuscript are available at \url{https://github.com/gammarama/intRo}.
